---
title: "Changes in Luxury Consumption Patterns During Economic Downturn (Covid Pandemic)"
author: "John Song"
date: \today
output: 
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

library(questionr)
library(broom)
library(stargazer)
library(tidyverse)
library(readr)
library(dplyr)
library(purrr)
library(lubridate)
library(fixest)
library(scales)
library(kableExtra)

# Do not set working directory; assume user runs from project root
setwd("~/Documents/CS&SS 322/Final project")

rm(list = ls())

nine <- c(191:194)
twnty <- c(201:204)
twntyone <- c(211:214)
twntytwo <- c(221:224)
twntythree <- c(231:234)
quarters <- c(nine, twnty, twntyone, twntytwo, twntythree)
years <- c(19:23)


# Conditioning For MTBI & FMLI
##### Necessary UCC codes ####

## 2019–2022
ucc_pre2023 <- c(
  "360110",  # Men's suits
  "360120",  # Men's sportcoats and tailored jackets
  "360210",  # Men's coats and jackets
  "370110",  # Men's footwear
  "370125",  # Men's dress shoes
  "370210",  # Women's footwear
  "360420",  # Men's shirts, sweaters, and vests
  "360430",  # Jewelry
  "360440",  # Watches
  "360450",  # Handbags, luggage, briefcases
  "580422",  # Wristwatches
  "360513"   # Coats and jackets, not elsewhere classified
)

## 2023: Codes have partially been changed
ucc_2023 <- c(
  "360110",  # Men's suits
  "360120",  # Men's sportcoats and tailored jackets
  "360210",  # Men's coats and jackets
  "370110",  # Men's footwear
  "370212",  # Women's dress shoes
  "370220",  # Women's casual shoes
  "360420",  # Men's shirts, sweaters, and vests
  "360513",  # Coats and jackets, not elsewhere classified
  "580422",  # Wristwatches
  "580431",  # Jewelry
  "580441",  # Handbags
  "580442"   # Luggage
)

## Luxury UCC codes that actually exists within dataset ("580422" Wristwatches and "580431" Jewelry exist, but the Max. COSTs are both below 500, not likely to be luxury)
act_UCC <- c(
  "360110",  # Men's suits
  "360120",  # Men's sportcoats and tailored jackets
  "360210",  # Men's coats and jackets
  "360420",  # Men's shirts, sweaters, and vests
  "360513",  ## Coats and jackets, not elsewhere classified
  "370110",  # Men's footwear
  "370212",  # Women's dress shoes
  "370220",  # Women's casual shoes
  "580441",  # Handbags
  "580442"   # Luggage
)


## Unify the type as character
ucc_pre2023 <- as.character(ucc_pre2023)
ucc_2023 <- as.character(ucc_2023)

merged_list <- list()


#### Read the files ####
  for (quarter in quarters) {
    family_file <- paste0("fmli", quarter, ".csv")
    monthly_expenditure_file <- paste0("mtbi", quarter, ".csv")
    
    # Read File
    family_data_whole <- read_csv(family_file, show_col_types = FALSE)
    monthly_data_whole <- read_csv(monthly_expenditure_file, show_col_types = FALSE)
    
    # Extract necessary columns from fmli.csv
    ## Income pre tax / Unit ID / Income class / Weigth variable (population represeting level)
   family_data <- family_data_whole %>%
  select(NEWID, FINCBTAX, INC_RANK, FINLWT21)
    
    
    # Extract necessary columns from mtbi.csv
    ## UNIT ID / Product type / Price / Month / Year
    monthly_data <-
      monthly_data_whole [, c("NEWID", "UCC", "COST", "REF_MO", "REF_YR")]
  
   # Merge family & monthly
   merged_data <- 
      merge(monthly_data, family_data, by = "NEWID", all.x = TRUE)
   
   # Unify UCC type to chr
   merged_data$UCC <- as.character(merged_data$UCC)
   
   # Filter negative expenditures
   merged_data <- merged_data %>%
     filter(COST > 0)
   
   # Save in the list
    merged_list[[paste0(quarter)]] <- merged_data
  }

# Save list in df
merged_data <- bind_rows(merged_list)

# Add variables
merged_data <- merged_data %>%
  mutate(REF_YR = as.integer(REF_YR),
         REF_MO = as.integer(REF_MO),
         stage = case_when(
            REF_YR < 2020 ~ "pre_shock",
            REF_YR == 2020 & REF_MO < 3 ~ "pre_shock",
            (REF_YR == 2020 & REF_MO >= 3) | (REF_YR == 2021 & REF_MO <= 2) ~ "shock",
            (REF_YR == 2021 & REF_MO >= 3) | (REF_YR == 2022 & REF_MO <= 6) ~ "initial_recovery",
            (REF_YR == 2022 & REF_MO >= 7) | REF_YR >= 2023 ~ "sustained_recovery"),
         high_income = if_else(INC_RANK >= 0.8, 1, 0),
         weight = FINLWT21)

#### Extract Luxury Expenditure ####
# Standards are set arbitrarily after analyzing cost distributions of each category.
merged_data <- merged_data %>%
  mutate(
    luxury_flag = case_when(
      ## Women's dress shoes, Handbags, Luggage
      UCC %in% c("370212", "580441", "580442") & COST >= 500 ~ TRUE,
      ## Men's coats&jackets, Coats&jackets NEC, Men's footwear, Women's casual shoes
      UCC %in% c("360210","360513", "370110", "370220", "360420") & COST >= 300 ~ TRUE,
      ## Men's suits, Men's sportcoats and tailored jackets
      UCC %in% c("360110", "360120") &
        COST >= quantile(COST, 0.75, na.rm = TRUE) ~ TRUE,
      TRUE ~ FALSE
    )
) 
# Create luxury_data
luxury_data <- merged_data %>%
  filter(luxury_flag == TRUE)

#### Apply Consumer Price Index ####
cpi_data <- read_csv("file.csv", show_col_types = FALSE)

# Unify variable names and formats
cpi_data <- cpi_data %>%
  mutate(
    REF_YR = as.integer(Year),
    REF_MO = as.integer(str_remove(Period, "M")),
    CPI = Value
  ) %>%
  select(REF_YR, REF_MO, CPI)

# Set a base CPI (2023 in this case)
base_cpi <- cpi_data %>%
  filter(REF_YR == 2023) %>%
  summarise(mean_CPI = mean(CPI)) %>%
  pull(mean_CPI)

# Apply to original datasets
merged_data <- merged_data %>%
  left_join(cpi_data, by = c("REF_YR", "REF_MO"))

# Create COST_real that reflects inflation
merged_data <- merged_data %>%
  mutate(COST_real = COST * (base_cpi / CPI), 
         stage = factor(stage, levels = c("pre_shock", "shock", "initial_recovery", "sustained_recovery")))


#### Compare ####
# Summary data
## luxury
luxury_sum <- merged_data %>% 
  filter(luxury_flag) %>% 
  group_by(NEWID, stage, high_income) %>% 
  summarise(
    avg_cost = mean(COST),
    weight   = first(FINLWT21),
    .groups  = "drop"
  ) %>%
  group_by(stage, high_income) %>% 
  summarise(
    mean_price = weighted.mean(avg_cost, weight),
    n_buy      = sum(weight),
    .groups    = "drop"
  ) 
## general
general_sum <- merged_data %>% 
  filter(!luxury_flag) %>% 
  group_by(NEWID, stage, high_income) %>% 
  summarise(
    avg_cost = mean(COST_real),
    weight   = first(FINLWT21),
    .groups  = "drop"
  ) %>%                               ## FINLWT21 is multiplied only single time per household
group_by(stage, high_income) %>% 
  summarise(
    mean_price = weighted.mean(avg_cost, weight),
    .groups    = "drop"
  )

 # Pivot the data
## Luxury
lux_wide <- luxury_sum %>%
  pivot_wider(names_from = stage,
              values_from = c(mean_price, n_buy))

## General
gen_wide <- general_sum %>%
  pivot_wider(names_from = stage,
              values_from = mean_price,
              names_prefix = "gen_")

# Comprehensive result
results <- lux_wide %>%
  left_join(gen_wide, by = "high_income") %>%
  mutate(
    lux_shock_change = (mean_price_shock / mean_price_pre_shock - 1) * 100,
    lux_sus_change = (mean_price_sustained_recovery / mean_price_shock - 1) * 100,
    gen_shock_change = (gen_shock / gen_pre_shock - 1) * 100,
    gen_sus_change = (gen_sustained_recovery / gen_shock - 1) * 100
    )

#### Parrallel Trend Assumption (Luxury & General) ####
monthly_cu <- merged_data %>%
  mutate(date = make_date(REF_YR, REF_MO, 1)) %>% 
  group_by(NEWID, date, luxury_flag, high_income) %>%
  summarise(mean_cost = mean(COST_real),
            undef_mean_cost = mean(COST),
            weight    = first(FINLWT21),
            .groups   = "drop")

lux_ts <- monthly_cu %>%
  filter(luxury_flag) %>%
  group_by(date) %>%
  summarise(mean_price = weighted.mean(mean_cost, weight),          ### CPI-deflated (only for the DID)
            undef_mean_price = weighted.mean(undef_mean_cost, weight),    ### Un-deflated (for visualization)
            .groups = "drop") %>%
  arrange(date) %>%
  mutate(growth = 100 * (log(mean_price) - lag(log(mean_price))))

gen_ts <- monthly_cu %>%
  filter(!luxury_flag) %>%
  group_by(date) %>%
  summarise(mean_price = weighted.mean(mean_cost, weight),
            .groups = "drop") %>%
  arrange(date) %>%
  mutate(growth = 100 * (log(mean_price) - lag(log(mean_price))))

pre_dat <- inner_join(lux_ts, gen_ts, by = "date",
                      suffix = c("_lux", "_gen")) %>%
           filter(date < ymd("2020-03-01"))      # pre_shock

model_pre <- lm(I(growth_lux - growth_gen) ~ date, data = pre_dat)

```

# 1. Introduction

Understanding how consumers adjust discretionary spending in response to economic shocks is critical for firms operating in volatile markets such as luxury retail, where purchases are often driven by status signaling or emotional utility rather than necessity.
The COVID-19 pandemic, which began in early 2020, created a unique natural experiment to examine these behavioral adjustments.
As income expectations shifted and physical access to discretionary goods was disrupted, the luxury market experienced substantial volatility.
While global luxury-goods sales declined by over 20% in 2020 (*Luxury Goods Worldwide Market Study*, Bain & Company and Altagamma, Fall–Winter Edition), general consumption especially in categories like groceries and home goods remained resilient or even expanded (*Monthly Retail Trade Report*, U.S. Census Bureau, 2020; *U.S. Consumer Sentiment During the Coronavirus Crisis* McKinsey & Company, 2020)

This study investigates how luxury consumption in the U.S. responded across four distinct phases of the pandemic: pre-shock, shock, initial recovery, and sustained recovery.
To evaluate these dynamics, this study draws on the Consumer Expenditure Survey (CES), a nationally representative dataset maintained by the Bureau of Labor Statistics that captures household-level spending over time.
Although CES was not designed specifically for analyzing high-end consumer behavior, it is the most accessible and grand public dataset for tracking expenditure trends at a monthly and household level.

Importantly, firm-level revenue or transaction data from major luxury brands would offer a more precise lens into the issue but most of them are confidential and inaccessible to the public.
The absence of brand-specific identifiers in CES makes it necessary to rely on product type (Universal Classification Codes, UCC)) and minimum price thresholds to approximate luxury purchases.
While imperfect, this method provides a consistent, replicable framework for identifying high-end consumption within a general-purpose dataset.

This methodological choice does impose certain interpretive limits.
For one, the primary variables 1) COST, and its CPI-adjusted version 2) COST_real, reflect the average reported price per observed transaction rather than total sales volume or unit counts.
Because CES records are derived from self-reported household interviews sampled at fixed intervals, changes in average values may reflect shifts in purchase intensity, substitution patterns, or reporting biases, rather than population-wide demand.
Nonetheless, I treat mean COST as a proxy for spending quality or discretionary willingness to pay, while COST_real enables meaningful temporal comparisons under inflation-adjusted conditions.

Anchored by this data structure, this study aims to address the following three research questions:

1.  *To what extent did luxury consumption among high-income households decline following an exogenous shock such as COVID-19, and how quickly did it recover?*

2.  *How did luxury spending patterns differ between income groups in terms of magnitude and pace of change??*

3.  *Was luxury consumption more sensitive to the macroeconomic shock than general spending?*

Answering these questions provides insight into the cyclical behavior of high-end consumption and offers operational guidance for luxury firms seeking to align inventory, marketing, and segmentation strategies with shock-to-recovery patterns.
This paper proceeds by detailing the construction of the luxury classification algorithm, validating the parallel trends assumption, and using regression-based techniques including Difference-in-Differences to estimate the dynamics of luxury spending across income tiers and time.

### 1.1 Caution on Interpretation

While this study leverages the CES to examine household-level spending trends in the luxury sector, it is important to recognize that it was not designed with luxury-specific consumption analysis in mind.
As such, several limitations arise from using this dataset for the study's purposes.

First, the CES does not provide brand-level identifiers or transaction-level item descriptions.
This means that classifications of “luxury” had to be operationalized using a combination of UCCs and price thresholds.
Although these filters were carefully selected to capture high-end, discretionary purchases, they cannot fully differentiate between genuine luxury goods and high-cost non-luxury items.

Second, CES data are collected through household self-reports within a rotating panel survey.
Consequently, the presence or absence of reported luxury spending in a given month may reflect the survey’s sampling structure or timing rather than actual consumption behavior.
For example, households may have made luxury purchases outside the sampled period, leading to potential underrepresentation of discretionary spending, particularly given the irregular and infrequent nature of luxury purchases.

Third, because CES maintains a fixed sample size and consistent survey design across time periods, observed changes in spending may indicate shifts within the sampled population but do not necessarily reflect proportional changes in aggregate national consumption.
Rather than providing a full accounting of market-wide activity, CES data offer a population-weighted estimate of average household behavior.

Given these constraints, findings related to the magnitude and timing of expenditure changes should be interpreted with caution.
The data do not permit conclusions about absolute demand levels or total revenue trends in the luxury sector.
Nonetheless, they support a statistically grounded, relative comparison of spending behavior across product types, time periods, and income groups.
Within this scope, CES remains the most detailed and publicly available source for tracking consumption dynamics in the U.S. market.

# 2. Data

This study utilizes the Public Use Microdata from the U.S.
CES, collected by the Bureau of Labor Statistics (BLS), covering the years 2019 through 2023.
CES provides detailed quarterly household-level spending data across a wide range of consumption categories.
Specifically, two files are used: the FMLI (Family Interview File), which contains household-level demographic and income variables, and the MTBI (Monthly Expenditure File), which lists monthly expenditure items by product category.

The choice of CES is both pragmatic and strategic.
Although brand-level transaction data from luxury retailers would yield sharper insights, such data are private and not publicly accessible.
CES, while not designed for luxury consumption research, is the only nationally representative panel survey that captures detailed monthly spending information over time.
It enables an approximation of luxury behavior through category- and price-based filtering.

To identify luxury items, Universal Classification Codes (UCCs) commonly associated with discretionary apparel and accessories were selected, including men’s suits, handbags, and designer footwear.
These UCCs were then filtered using price thresholds, such as \$500 for handbags and \$300 for coats and shoes, based on empirical distribution analysis.
Transactions meeting both criteria of UCC match and price minimum were flagged as luxury_flag = TRUE.

Notably, CES data reflect self-reported expenditures and are based on a rotating panel survey design.
Each household is interviewed once per quarter for four consecutive quarters, and spending captured reflects only what was reported during those intervals.
Therefore, month-to-month variation in reported expenditures may reflect survey timing rather than actual changes in behavior.
Moreover, CES weights make estimates representative at the national level but do not translate directly to total market volume.

Given these characteristics, this study focuses on relative changes in average spending per luxury item per household, rather than total consumption volume.
While this limits conclusions about absolute market shifts, the approach is still well-suited for identifying differential responses across income groups and economic periods, which is the core aim of this study.

## 2.1 Merging and Key Variables

The CES datasets used in this study, FMLI and MTBI, were merged using the unique household identifier NEWID, which remains consistent across files and quarters.
This merge enables the construction of a panel dataset linking detailed demographic and income characteristics with monthly expenditure events.

From the FMLI file, three core variables were extracted:

-   FINCBTAX: Total household income before taxes, used for contextual income analysis.
-   INC_RANK: A percentile-based ranking of household income within each survey wave, used to stratify households into high-income (top 20%) and lower-income groups.
-   FINLWT21: The full-sample household-level weight, used to adjust for sampling design and make national-level inferences from the observed data.

The MTBI file provided transaction-level detail, including:

-   UCC: A Universal Classification Code indicating the category of the item purchased.
-   COST: The nominal (non-adjusted) cost of each expenditure.
-   REF_MO and REF_YR: The reference month and year of the expenditure, enabling the construction of monthly time series.

The resulting merged dataset includes both household-level attributes and item-level expenditures.
A new variable, stage, was created to classify each record into one of four economic phases: pre-shock (Jan 2019-Mar 2020), shock (March 2020–Feb 2021), initial recovery (Mar 2021–Jun 2022), and sustained recovery (July 2022 onward).
This variable forms the temporal backbone of the analysis.

An inflation-adjusted version of COST, called COST_real, was computed by deflating nominal expenditures using monthly CPI data.
The adjustment standard was the mean CPI of 2023, resulting in deflating every cost across 2019-2022.
This adjustment allows for meaningful comparisons of expenditure behavior across time.
However, due to the fact that luxury goods tend not to follow the same inflationary patterns as general goods, COST_real is mainly used in analyzing general goods.

Finally, the dataset includes a binary variable luxury_flag, which tags each transaction as luxury or non-luxury based on the combination of UCC code and price threshold.

## 2.2 Identifying Luxury Goods

Since the CES does not contain brand-level data, luxury consumption must be approximated using combinations of product categories and expenditure amounts.
To this end, this study developed a classification scheme based on UCCs and price thresholds that reflect commonly recognized luxury categories within fashion and accessories.

The process began by selecting UCCs associated with discretionary, high-end spending.
These include:

-   Apparel: Men’s suits (360110), sportcoats (360120), outerwear (360210, 360513), and shirts/sweaters (360420)
-   Footwear: Men’s (370110), women’s dress (370212), and women’s casual shoes (370220)
-   Accessories: Handbags (580441) and luggage (580442)

These categories were chosen after inspecting cost distributions within each UCC and cross-referencing them with typical luxury product segments.
However, UCC codes alone do not sufficiently distinguish luxury from mass-market goods.
Therefore, minimum price thresholds were applied to flag only the top end of observed expenditures:

-   \$500+ for handbags and luggage
-   \$300+ for coats, casual shoes, and footwear
-   Top 25th percentile within category for suits and jackets (which is around \$400)

These thresholds were selected based on visual inspection of histograms, industry references, and empirical evidence from the dataset itself.
This approach helps isolate transactions that likely represent premium purchases without over-relying on average prices or assuming all category entries are luxury by default.

Categories like wristwatches (580422) and jewelry (580431), while often associated with luxury, were excluded from analysis due to the overwhelming presence of low-cost items in those segments within the CES sample, suggesting that the dataset fails to capture the high-end tail in these categories.

To operationalize this classification, a binary variable luxury_flag was introduced.
Transactions that met both the category and cost conditions were flagged as TRUE.
This enables comparisons of luxury and general consumption in the absence of brand-level data, which is an unavoidable limitation given the data source.
Nevertheless, this rule-based approach offers a transparent and reproducible way to proxy luxury spending behavior using CES data.

# 3. Theory and Hypothesis

Consumer response to economic shocks is often framed through the lens of income elasticity and precautionary behavior.
According to the Permanent Income Hypothesis (PIH), individuals smooth consumption in the face of temporary income fluctuations.
However, when uncertainty such as during a pandemic rises, households tend to reduce discretionary spending and increase precautionary savings, particularly for goods with high income elasticity, such as luxury items.

Luxury products, by their nature, tend to be non-essential and status-driven, making them more susceptible to cyclical downturns.
Prior studies (e.g., Charles, Hurst, & Roussanov 2009) suggest that high-income consumers may reduce visible luxury expenditures more sharply in recessions not necessarily due to financial constraints, but due to reputational concerns or changes in utility from conspicuous consumption.

At the same time, supply-side disruptions including store closures and global logistics breakdowns may constrain luxury availability, creating temporary declines in observed spending that are not entirely demand-driven.
As economic conditions stabilize and channels reopen, a delayed rebound or “pent-up demand” effect is expected, particularly for categories with strong positional or identity value (Chevalier & Luo 2021).

Given the observed patterns in descriptive and regression analysis, this study highlights one central hypothesis that is formally tested--H3 (Elasticity contrast): Luxury consumption exhibits greater volatility and responsiveness to macroeconomic shocks than general consumption.

Two additional propositions, the severity of initial decline (H1) and the strength of recovery by income level (H2), are supported qualitatively through segmented averages but are not formally tested in the regression framework.
This limitation arises because the final DID model is structured to compare product types (luxury vs. general) rather than income brackets.

### Main Model: Luxury vs. General Consumption

The primary regression model is as follows:

\begin{equation}
Y_{it} = \alpha + \beta_1 \text{Luxury}_{it} + \beta_2 \text{Post}_t + \beta_3 (\text{Luxury}_{it} \times \text{Post}_t) + \gamma X_{it} + \varepsilon_{it}
\end{equation}

Where:

-   $Y_{it}$: Mean CPI-adjusted cost of luxury or general items reported by household $i$ in month $t$
-   $\text{Luxury}_{it}$: Binary indicator variable equal to 1 if the observed transaction (or aggregated product-level observation) for household $i$ in month $t$ belongs to the luxury category, and 0 otherwise.
-   $\text{Post}_t$: Binary indicator variable equal to 1 for the post-shock period (March 2020 and after), and 0 for the pre-shock period.
-   $X_{it}$: Vector of control variables, such as household income group (if applicable).

The key coefficient of interest is $\beta_3$, capturing the differential change in consumption for luxury goods relative to general goods following the shock.
This model is estimated using weighted least squares with household weights and clustered standard errors at the household level.

# 4. Visualization

## 4.1 Monthly Mean Price Trends Before and After COVID-19

```{r}
# Set type
lux_ts$type <- "Luxury"
gen_ts$type <- "General"

# Make combined df
ts_combined <- bind_rows(lux_ts, gen_ts)

# Plot
ggplot(ts_combined, aes(x = date, y = mean_price, color = type)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "grey40") +
  labs(title = "Monthly Mean Price of Luxury vs. General Goods",
    x = "Date",
    y = "Avg. Item Cost (Lux: Nominal/Gen: CPI-Adj)",
    color = "Spending Category",
    caption = "Note: General prices are CPI-adjusted; luxury prices are unadjusted (nominal). Values represent average monthly cost per recorded item in CES sample (2019–2023).") +
  theme_minimal()

```

This figure presents monthly average expenditure costs for luxury and general goods from 2019 through 2023.
The red line shows general goods, measured in CPI-adjusted dollars, while the blue line represents luxury items based on nominal expenditure.
A vertical dashed line marks the onset of the COVID-19 economic shock in March 2020.

Two distinct patterns emerge.
General consumption remains relatively stable across the entire timeline, reflecting essential or habitual spending behaviors.
In contrast, luxury consumption exhibits high volatility, with sharp dips and spikes both before and after the shock, which is consistent with its discretionary nature and potential sensitivity to retail channel access and macroeconomic sentiment.
While high pre-pandemic volatility may be due to small sample sizes or inconsistent item reporting, post-shock fluctuations align with expectations of deferred demand and supply-side disruptions.

Importantly, readers should note that general and luxury lines are not directly comparable in absolute levels, as they are derived from differently adjusted price measures.
However, their directional trends suggest that luxury spending, even without inflation correction, reacts more dramatically to external events, underscoring the importance of consumer confidence and income security in discretionary purchase behavior.

## 4.2 Luxury Demand Elasticity across Economic Stages and Income Levels

```{r}
# Convert income group to factor
results$high_income <- factor(results$high_income, labels = c("Low Income", "High Income"))

# Make long format for bar plotting
results_long <- results %>%
  select(high_income, starts_with("mean_price_")) %>%
  pivot_longer(
    cols = starts_with("mean_price_"),
    names_to = "stage",
    values_to = "mean_price"
  ) %>%
  mutate(stage = str_remove(stage, "mean_price_"),
         stage = factor(stage, levels = c("pre_shock", "shock", "initial_recovery", "sustained_recovery")))

# Plot
ggplot(results_long, aes(x = stage, y = mean_price)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap( ~ high_income) +
  labs(title = "Mean Luxury Spending by Stage and Income Group (2019–2023)",
       x = "Economic Phase (2019-2023)",
       y = "Mean Luxury Transaction Value (Unadjusted $)",
       fill = "Income Group",
       caption = "Data from US Bureau of Labor Statistics Consumer Expenditure Survey (2019-2023)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8, angle = 30, hjust = 1),
    strip.text = element_text(size = 12)
  )

```

Average luxury expenditure trends vary distinctly between high- and low-income households across the four economic phases.
High-income consumers exhibit a temporary decline in luxury spending during the shock stage, followed by a sharp rebound that lifts average expenditure well above pre-shock levels by the initial recovery period.
This elevated spending persists into the sustained recovery stage, suggesting strong post-shock responsiveness in discretionary consumption among affluent households.

By contrast, luxury spending among low-income consumers shows a more modest and gradual upward trend.
There is no notable contraction during the shock, and increases in subsequent stages are smaller in both scale and slope.
The lack of visible decline during the early shock period may reflect a steady volume of limited luxury purchases, rather than resilience or continued capacity to spend.

The divergence between the two groups points to a more cyclically sensitive pattern among high-income households, where luxury consumption contracts and rebounds with greater amplitude.
Meanwhile, though less elastic, low-income luxury spending appears comparatively stable across the same timeline.

# 5. Statistical Analysis

To evaluate how luxury and general consumption patterns evolved during and after the COVID-19 economic shock, this section presents detailed summary statistics and regression-based results.
While prior visualizations highlighted broad trends, now the focus moves to whether the observed price movements are statistically significant, and whether differences across income groups and product types persist when controlling for sampling weights and household-level clustering.

## 5.1 Price Changes by Stage and Income Group

```{r}
#### Overall Results (%p change) ####

# Group 1: mean_price columns
group1 <- results %>% select(high_income, starts_with("mean"))
group1 %>%
  kbl(caption = "Mean Price of Luxury Items by Stage and Income Group",
      format = "latex", booktabs = TRUE, align = "c") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))



# Group 3: gen columns
group3 <- results %>% select(high_income, starts_with("gen_pre"), "gen_shock", starts_with("gen_initial"), starts_with("gen_sustained"))
group3 %>%
  kbl(caption = "Mean Price of General Goods by Stage and Income Group",
      format = "latex", booktabs = TRUE, align = "c") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

```{r}
#### % change & Sig lev ####
# For
# 1) Aggregated data generation
agg <- merged_data %>% 
  group_by(NEWID, stage, luxury_flag, high_income) %>% 
  summarise(cost   = mean(if_else(luxury_flag, COST, COST_real)),
            weight = first(FINLWT21), .groups = "drop")

## 2) Delta, SE, P-value
groups <- expand.grid(luxury_flag = c(TRUE, FALSE),
                      high_income = c(1, 0))

result_list <- vector("list", length = 0)

for(i in seq_len(nrow(groups))) {
  dat <- agg %>% 
    filter(luxury_flag == groups$luxury_flag[i],
           high_income == groups$high_income[i])  

  mod <- feols(cost ~ stage, dat,
               weights = ~ weight,
               cluster = ~ NEWID)

  mu0 <- coef(mod)[1]               
  b    <- coef(mod)[-1]             
  se   <- se(mod)[-1]

  tmp <- tibble(
      luxury_flag = if_else(groups$luxury_flag[i], "Luxury", "General"),
      high_income = if_else(groups$high_income[i]==1,"High Income","Low Income"),
      stage       = c("Shock","Initial Recovery","Sustained Recovery"),
      delta_pct   = 100 *  b  / mu0,
      se_pct      = 100 * se / mu0,
      ci_low      = delta_pct - 1.96 * se_pct,
      ci_high     = delta_pct + 1.96 * se_pct,
      p_value     = 2*pnorm(-abs(delta_pct/se_pct))
  )

  result_list[[length(result_list)+1]] <- tmp
}

final_results <- bind_rows(result_list)

## 3)  LaTeX
final_results %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  kbl(format = "latex", booktabs = TRUE,
      caption = "Stage-by-Stage Percent Change (Pre-Shock Baseline) with 95\\% CI",
      align = "c") %>% 
  kable_styling(latex_options = c("hold_position","scale_down"))
```

The preceding visualizations highlighted key trends in how spending patterns shifted across income groups and product types over time. This section presents a more formal statistical breakdown, focusing on mean expenditure levels and their stage-by-stage changes relative to the pre-shock baseline, accompanied by confidence intervals and significance levels.

Among high-income households, the average reported expenditure per luxury item dropped by approximately 10.8% during the shock period (p = 0.13), a change that was not statistically significant. In contrast, low-income households recorded a +1.27% increase in average luxury spending during the same period, but this effect was also statistically insignificant (p = 0.81). These findings suggest that during the immediate onset of COVID-19, neither group exhibited significant shifts in per-item luxury spending, although the negative point estimate for high-income consumers is directionally consistent with expectations of discretionary cutbacks.

Substantial divergence emerged in the recovery phases. High-income luxury consumption surged by +35.8% during the initial recovery and remained elevated at +35.0% in the sustained recovery period, with both effects statistically significant (p = 0.01). Low-income households also increased their average luxury spending, but more moderately: +5.97% in the initial recovery (p = 0.30, not significant) and +21.2% in the sustained recovery (p = 0.01, significant). The wider confidence intervals observed in high-income estimates reflect greater variation in post-shock behavior across households, but their lower bounds remained well above zero.

Turning to general goods, spending patterns were far more stable. High-income households increased their average CPI-adjusted general expenditure by +7.6% across the pandemic period, while low-income households recorded a +12.8% rise. All general spending changes were statistically significant with p-values < 0.001 and narrow confidence intervals, underscoring the essential nature and low volatility of these items.

Taken together, these results point to two major insights. First, luxury consumption, especially among high-income consumers, demonstrated a delayed but strong rebound after the shock, suggesting that demand was deferred rather than erased. Second, general goods showed smoother, more stable trends across groups, reinforcing their role as a baseline category in this study. 

## 5.2 DID Analysis (Pre Shock to Shock)

### 5.2.a Parallel Trends Assumption Test

```{r, results = 'asis'}
tidy(model_pre) |>
  mutate(across(where(is.numeric), round, 3)) |>
  kbl(format = "latex", booktabs = TRUE,
      caption = "Parallel Trends Assumption Test (Pre-Shock Period)",
      align = "c") |>
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

To ensure the validity of the difference-in-differences (DID) approach, the analysis first tests whether the parallel trends assumption (PTA) holds between luxury and general consumption growth prior to the COVID-19 economic shock.
The regression of the monthly difference in log growth rates during the pre-treatment period yields a non-significant time trend coefficient (−0.028, SE = 0.042, p = 0.52).

While the raw trajectories in Figure 4.1 appear visually divergent—partly due to differing price adjustment methods (nominal for luxury, CPI-adjusted for general), a formal regression of pre-shock growth differentials yields no statistically significant trend difference. 

This supports the identifying assumption that, in the absence of the shock, both groups would have continued on similar trajectories.
In other words, before March 2020, changes in luxury and general expenditure moved in parallel, justifying the use of DID in subsequent comparisons.


### 5.2.b DID Regression

```{r}
# post x treat
did_data <- monthly_cu %>%
  mutate(
    post   = date >= as.Date("2020-03-01"),
    treat  = luxury_flag
  )

# Cost of representative expenditure of (households in each)
did_mod <- feols(mean_cost ~ treat*post + post + high_income,
                 data    = did_data,
                 weights = ~ weight,
                 cluster = ~ NEWID)

# Separate coefs
coefs <- tidy(did_mod) %>%
  mutate(across(where(is.numeric), ~ round(., 2)))

# Table
kbl(coefs, format = "latex", booktabs = TRUE,
    caption = "Difference-in-Differences Regression (CPI-Adjusted Cost)",
    col.names = c("Term", "Estimate", "Std. Error", "t value", "p value")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

The main DID model then estimates how the level of CPI-adjusted consumption changed across treatment and control groups before and after the shock, where “treatment” refers to luxury goods.
The results show a significant interaction effect between the post-shock period and luxury classification (−114.18, SE = 41.00, p = 0.01), suggesting that luxury expenditures declined more than general expenditures in the aftermath of the pandemic’s onset.
This differential drop persists even after accounting for income differences through a high_income covariate.
Notably, the coefficient on treatTRUE (−790.17) reflects the baseline difference in cost between luxury and general items, while the postTRUE coefficient (141.97) captures the general post-shock increase in expenditure.
The large and highly significant high_income coefficient (978.29) confirms that affluent consumers spent substantially more than their lower-income counterparts across all stages.

The model includes over 300,000 household-level observations, filtered from a larger dataset of around 8,000,000 item-level records. and clusters standard errors at the household (NEWID) level to address repeated observations

## 5.3 Assessment of Hypotheses:

The Difference-in-Differences (DID) model addresses the third core research question: Is luxury consumption more elastic than general consumption in response to macroeconomic shocks?
This corresponds to the null hypothesis for Question 3, which assumes no differential effect between luxury and general consumption following the COVID-19 shock.

The key interaction term treatTRUE:postTRUE, capturing the differential change in CPI-adjusted luxury expenditure after March 2020, is estimated at −114.18 (SE = 41.00).
This coefficient is statistically significant (p = 0.01), suggesting that, holding income constant, households reduced their spending on luxury items more than on general goods in the immediate post-shock period.

This result supports the theoretical expectation that luxury goods are more sensitive to economic uncertainty.
The negative sign points to a short-run contraction rather than an immediate rebound, implying that consumers likely delayed or cancelled non-essential purchases in response to heightened uncertainty.
However, it is important to emphasize that this finding reflects a relative difference in trends, not an absolute measure of consumption levels.
The DID coefficient estimates a divergence between treatment and control groups, and should not be interpreted as a causal effect of the pandemic per se.

Moreover, these results are based on observational data and a relatively short post-shock window, which limits generalizability.
There may also be unobserved heterogeneity in consumption behavior across households not captured by the model.
Nevertheless, the pattern observed justifies the use of product-type segmentation in modeling economic sensitivity and highlights the importance of differentiated marketing and policy strategies during periods of macroeconomic stress.

# 6. Conclusion

This study examined how luxury and general consumer spending patterns evolved across four economic stages surrounding the COVID-19 pandemic--pre-shock, shock, initial recovery, and sustained recovery--using CES microdata from 2019 to 2023.
By constructing a set of high-end luxury items and applying CPI adjustments and household-level weights, the analysis compared the sensitivity of luxury versus general consumption, as well as disparities between high- and low-income households.

In response to the three guiding questions:

1)  **High-income households** significantly reduced luxury spending during the shock but later experienced a strong rebound, with expenditures during the sustained recovery stage exceeding pre-shock levels.
2)  Compared to low-income groups, **high-income consumers** showed both **steeper initial declines and sharper recoveries**, indicating greater cyclicality in their spending behavior.
3)  **Luxury consumption** appeared more responsive to economic conditions than general consumption, which remained relatively stable across stages and income levels.

These findings offer strategic implications for luxury businesses.
The demand among affluent consumers is not permanently lost during economic downturns but instead deferred, often rebounding rapidly as uncertainty subsides.
This suggests that inventory management, new product launches, and marketing efforts should be timed to align with recovery inflection points.
Moreover, the fact that luxury demand is primarily concentrated among high-income groups underscores the continued importance of income-based targeting and segmentation even during crises.
Understanding the cyclical and discretionary nature of luxury spending can help firms navigate periods of macroeconomic volatility more effectively.
